---
title: "Annotation_report"
author: "Miguel Tripp"
date: "2020-11-23"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

# TRINOTATE Annotation report

-   We used the package TrinotateR to explore Annotation results from Trinotate.

-   The used assembly was the filtered with **TPM** and **CD-HIT**

## Filter Non-Eukariote sequences

```{r Trinotate R, message=FALSE, warning=FALSE}
library(trinotateR)
library(gplots)
library(knitr)
library(tidyverse)
library(here)

file_good = here("data", "Annotation", "Trinotate_report.xls")
x_good <- read_trinotate(file_good)

#summary table
kable(summary_trinotate(x_good), caption = "Table 1: summary table for all annotated transcripts")

#filter only eukarioye sequences in Blastx
x_good_nonEu <- x_good %>%
  filter(!str_detect(sprot_Top_BLASTP_hit, "Eukaryota"))

#Thereafter, with filter-out the non Eucaryote sequences

x_good <- x_good[!x_good$transcript_id %in% x_good_nonEu$transcript_id, ]

kable(summary_trinotate(x_good), caption = "Table 2: summary table for only eukaryote transcritps as on BlastP")

```

## Extract each of the Trinotate report elements

```{r message= FALSE}
#pfam <- split_pfam(x_good)
#spfam <- summary_pfam(pfam)
#We don´t need pfam in this paper

go <- split_GO(x_good, hit = "gene_ontology_blast")
gos <- summary_GO(go)

blastx <- split_blast(x_good, "sprot_Top_BLASTX_hit") 
sblastx <- summary_blast(blastx)

blastp <- split_blast(x_good, "sprot_Top_BLASTP_hit")
sblastp <- summary_blast(blastp)


#we create a blastp_tophit to reteain only the hit with the lowest evalue
blastp_topHit <- blastp %>%
   dplyr::group_by(transcript) %>%
  dplyr::slice(which.min(evalue)) %>%
  ungroup 

```

## Taxonomy plot based on results from nucleotides and ORFs (Bastx and Blastp)

```{r taxonomy,echo= FALSE}
# ====== Taxonomic classification on BlastP top hit hits=======
# only based on genus
genus.p <- blastp %>% 
  count(Genus = fct_lump(genus, n = 35)) %>% 
  arrange(desc(n))


# ====== Taxonomic classification on BlastX top hit hits=======
# only based on genus
genus.x <- blastx %>% 
  count(Genus = fct_lump(genus, n = 35)) %>% 
  arrange(desc(n))


#####combined plot; blastx and blastp
genus.x$Type <-"Transcript"
genus.p$Type <-"ORF"

plot <- rbind(genus.p, genus.x)

ggplot(plot, aes(x = n, y = reorder(Genus,-n), fill = Type))+
  geom_bar(stat="identity")+
  theme_bw()+
  scale_fill_manual(values = c("lightblue", "darkcyan"))+
  #facet_grid(.~ Type)+
  labs(y = "Taxa", x = "Number of sequences") +
  theme(axis.text.y = element_text(hjust = 1, size = 7),
        axis.text.x = element_text(angle = 90),
        panel.grid = element_blank())

  
  
```

## Identity distribution of the annotated ORFs (BlastP)

```{r echo = FALSE}
iden_good <- ggplot(blastp, aes(x = identity))+
  geom_histogram(fill = "lightblue", alpha = 0.2)+
  geom_histogram(data = blastp_topHit, aes(x = identity), fill = "tomato", alpha = 0.2)+
  labs(x = "Identity (%)", y = "Number of sequences", title = "Identity distribution",
       caption = "The red histogram corresponds to Tophit BlastP result")
iden_good

```

## GO distribution

```{r, echo = FALSE}
library(ggpubr)
# ====
plotgos <- head(gos[order(-gos$transcripts),], 75)

gos_tid <- gos %>%
  arrange(desc(transcripts)) %>%
  dplyr::group_by(ontology) %>%
  dplyr::slice(1:20)

bar <- ggbarplot(gos_tid, "name", "transcripts",
                 fill = "ontology", 
                 color = "ontology",
                 xlab = "Name",
                 ylab = "Number of transcripts",
                 palette = c("#00AFBB", "#E7B800", "#FC4E07"),            # jco journal color palett. see ?ggpar
                 sort.val = "desc",           # Sort the value in dscending order
                 sort.by.groups = TRUE,
                 x.text.angle = 45,           # Rotate vertically x axis text
                 title = "Top 20 ontologies",
                 legend = "bottom",
                 orientation = "horizontal")+
  theme(axis.text = element_text(hjust = 1, size = 8))
bar

```
